---
title: "Geospatial Visualization of the Covid19 Cases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{geo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message=FALSE, 
  warning=FALSE,
  collapse = TRUE,
  fig.height=6, 
  fig.width=6, 
  comment = "#>"
)
```

The `covid19sf_geo` provides information about San Francisco Covid19 cases distribution by geospatial location. Also, testing locations across San Francisco available on `covid19sf_test_loc` dataset. The following vignette provides examples for geospatial visualization of those datasets. Both datasets are `sf` objects and contain geometric information (i.e., ready to plot on a map).

**Note:** This is a non-CRAN vignette, and the following libraries required to build the plots on this document:

```{r}
require(ggplot2)
require(mapview)
library(leafsync)
require(sf)
require(tmap)
require(RColorBrewer)
```

### The covid19sf_geo dataset

The `covid19sf_geo` dataset provides a snapshot of the distribution of the Covid19 cases in San Francisco by different geographic locations splits of the city. The dataset contains the following fields:

- `area_type` - the geograpichal split method:
  - `ZCTA` for view the data by ZIP code
  - `Analysis Neighborhood` for view the data by neigborhoods
  - `Census Tract` for view the data by census tract, and
  - `Citywide` for total cases in the city
- `id` - the area ID (e.g., ZIP code, neighborhood name, etc.)
- `count` - total number of positive cases in the area
- `rate` - cases rate per 10000 residents
- `deaths` - total number of deaths in the area
- `acs_population` - total number of residents in the area
- `last_updated` - most recent update time of the dataset


While the first three geographical split methods contain geometry components that enable us to plot them as a map, the last is just an aggregated summary of the city's total cases.

```{r }
library(covid19sf)

data(covid19sf_geo)

class(covid19sf_geo)

head(covid19sf_geo)
```

#### Plotting cases with mapview

The most intuitive method for plotting `sf` objects is with the [mapview](https://r-spatial.github.io/mapview/) package, which is a wrapper for the leaflet JavaScript package. The main advantage of the mapview package that it is both interactive and smoothly works with `sf` objects. The following example demonstrated the use case of the `mapview` function to plot the confirmed cases in San Francisco with the `plot` function to plot cases distribution by ZIP code:

```{r}
library(dplyr)

covid19sf_geo %>% 
  filter(area_type == "ZCTA") %>% 
  mapview(zcol = "count")
```



You can use `at` and `col.regions` arguments to define color buckets and color range, respectively:

```{r}
covid19sf_geo %>% 
  filter(area_type == "ZCTA") %>% 
  mapview(zcol = "count", 
          at = c(0,  200, 400, 800, 1200, 1600, 2000),
          col.regions = (c('#fef0d9','#b30000')))
```


#### Plotting vaccine data with tmap

The [tmap](https://r-tmap.github.io/tmap/index.html) package provides functions and tools for creating thematic maps. The package supports `sf` objects and follow the **ggplot2** framework. In the example below we will plot the COVID-19 vaccine data by geographic using the `covid19sf_vaccine_geo` dataset:

```{r}
data(covid19sf_vaccine_geo)

head(covid19sf_vaccine_geo)
```

**Additional setting:** Following changes in the default options of the **sf** package from version `1.0-1` by default option is to use s2 spherical geometry as default when coordinates are ellipsoidal. That cause some issues with the **tmap** package, therefore we will set this functionality as `FALSE`:

```{r}
sf_use_s2(FALSE)
```

We will use the `percent_pop_series_completed` variable to plot the percentage of population that finished their vaccination process. Let's first filter the data and transform the `percent_pop_series_completed` from decimal to percentage:

```{r}
df <- covid19sf_vaccine_geo %>% filter(area_type == "Analysis Neighborhood") %>%
  dplyr::mutate(perc_complated = percent_pop_series_completed * 100)
```

Now we can plot the new object:

```{r}
tm_shape(df) + 
  tm_polygons("perc_complated", 
              title = "% Group")
```

By default, the `tm_polygons` function bucket the numeric variable, in this case percentage of vaccinated population, into buckets. You can control the number of buckets using the `n` argument. Let's now start to customize the plot and modify to color palette:

```{r}
tm_shape(df) + 
  tm_polygons("perc_complated", 
              title = "% Group",
              palette = "RdYlBu")
```

We can customize the plot background with the `tm_style` function by using the `style` argument:

```{r}
tm_shape(df) + 
  tm_polygons("perc_complated", 
              title = "% Group",
              palette = "RdYlBu") +
  tm_style(style = "cobalt")
```

Last but not least, let's add title and labels for the geographic locations with the `tm_text` and `tm_layout` functions:


```{r}
tm_shape(df) + 
  tm_polygons("perc_complated", 
              title = "% Group",
              palette = "RdYlBu") +
  tm_style("cobalt") + 
  tm_text("id", size = 0.5) +
  tm_layout(
    legend.position=c("right", "bottom"),
    legend.outside = FALSE,
    legend.width = 1,
    legend.title.size = 1.2,
    legend.text.size = 1,
    # legend.outside.size = 0.9,
    title= paste("COVID-19 Vaccines Given", 
                 "to San Franciscans by Geography",
                 sep = " "), 
    title.position = c("left", "top") ,
    inner.margins = c(0.01, .01, .12, .25)) 
```


#### Plotting cases with base plot

The [sf](https://r-spatial.github.io/sf/index.html) package provides a `plot` method for `sf` objects (see `?sf:::plot.sf` for more information). Similarly to the previews examples above, we will replot the confirmed cases by ZIP code with the `plot` function:


```{r}


zip <- covid19sf_geo %>% 
  dplyr::filter(area_type == "ZCTA") %>% 
  dplyr::select(count, geometry) %>%
  plot(main = "Covid19 Cases by ZIP Code")

```

You can define the color palette with the `pal` argument and set the level of breaks of the color scale by setting the `breaks` argument to `quantile` and the number of breaks with the `nbreaks` argument (which should be aligned with the number of colors on the color palette):


```{r }
library(RColorBrewer)
pal <- brewer.pal(9, "OrRd")

covid19sf_geo %>% 
  filter(area_type == "ZCTA") %>% 
  select(count, geometry) %>%
  plot(main = "Covid19 Cases by ZIP Code",
       breaks = "quantile", nbreaks = 9,
       pal = pal)
```



#### Plotting cases with ggplot2

Plotting `sf` object can be done with the [ggplot2](https://ggplot2.tidyverse.org/) package natively by using the [geom_sf](https://ggplot2.tidyverse.org/reference/ggsf.html) function for plotting `sf` objects:

```{r}
library(ggplot2)

covid19sf_geo %>% 
  filter(area_type == "ZCTA") %>% 
  ggplot() + 
  geom_sf(aes(fill=count)) +
  ggtitle("Covid19 Cases by ZIP Code")
```

You can customize the polygon color scale by using the [scale_viridis](https://ggplot2.tidyverse.org/reference/scale_viridis.html) function that enables you to select different viridis color palettes. In addition, the [geom_sf_label](https://ggplot2.tidyverse.org/reference/ggsf.html) enables you to add labels for each polygon. In the next example, we will replot the count of cases by ZIP code, this time using `scale_fill_viridis_b` color palette and setting the `id` variable as the polygon title using the `geom_sf_label`:

```{r}
covid19sf_geo %>% 
  filter(area_type == "ZCTA") %>% 
  ggplot() + 
  geom_sf(aes(fill=count)) + 
  scale_fill_viridis_b() +
  geom_sf_label(aes(label = id)) + 
  ggtitle("Covid19 Cases by ZIP Code")
```


Additional customization of the viridis color palettes can be done by the `option` argument, where the `begin`, and `end` arguments control the color hue:

```{r}
covid19sf_geo %>% 
  filter(area_type == "ZCTA") %>% 
  ggplot() + 
  geom_sf(aes(fill=count)) + 
    scale_fill_viridis_b(option = "A",
                       begin = 0.2,
                       end = 0.7) + 
   theme_void() +
  ggtitle("Covid19 Cases by ZIP Code")
```


### The covid19sf_test_loc dataset

The `covid19sf_test_loc` datasets provides general metadata about the Covid19 testing locations in San Francisco:

```{r}
data(covid19sf_test_loc)

head(covid19sf_test_loc)
```

Plotting the testing locations on map is fairly similar to one of the `covid19sf_geo` as both are `sf` objects. The main distinction between the two, is that the `covid19sf_test_loc` provides the geometry location (e.g., latitude and longitude) as opposed to a polygon. Let's plot the locations with the **mapview** package setting the location color by the type (private or public):

```{r}
covid19sf_test_loc %>% mapview(zcol = "location_type")
```

### Combine cases dist. and testing points

The `sync` function from the **leafsync** package enables to combine multiple maps plots. In the following example, we will put side by side the cases split by ZIP code and the testing point in the city map:

```{r}

m1 <- covid19sf_geo %>% 
  filter(area_type == "ZCTA") %>% 
  mapview(zcol = "count", 
          at = c(0,  200, 400, 800, 1200, 1600, 2000),
          col.regions = (c('#fef0d9','#b30000')))
m2 <- covid19sf_test_loc %>% mapview(zcol = "location_type")
sync(m1, m2)
```


