---
title: "Hospital capacity data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hospital}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.height=5, fig.width=8,
   message=FALSE, warning=FALSE,
  collapse = TRUE,
  comment = "#>"
)
`%>%` <- magrittr::`%>%`
library(covid19sf)
```

The **covid19sf** package provides two datasets focusing on the overall hospital capacity and patients status in San Francisco hospitals:

* `covid19sf_hospital` - a daily snapshot of the hospital capacity
* `covid19sf_hospitalizations` - a daily snapshot of the patients status

### Hospital Capacity

The `covid19sf_hospital` dataset provides a daily snapshot of the San Francisco hospital capacity by bed type and status with the following variables:

* `bed_type` - number of hospital beds available by department type:
    * Acute Care
    * Intensive Care
    * Acute Care Surge
    * Intensive Care Surge
* `status` - the department corresponding bed status:
  * Available
  * COVID-19 (Confirmed & Suspected)
  * Other Patients


```{r}
data(covid19sf_hospital)

head(covid19sf_hospital)
```


Let's summarise the hospital status as of the most recent date available on the dataset:

```{r}
library(dplyr)

covid19sf_hospital %>%
  filter(date == max(date)) %>%
  group_by(bed_type, status) %>%
  summarise(total = sum(count))
```


```{r}
library(plotly)

covid19sf_hospital %>%
  filter(bed_type == "Acute Care") %>%
  mutate(status = factor(status, levels = c("Other Patients", "COVID-19 (Confirmed & Suspected)", "Available"))) %>%
  plot_ly(x = ~ date,
          y = ~ count,
          type = 'scatter', 
          mode = 'none', 
          color = ~status,
          stackgroup = 'one') %>%
    layout(title = "SF Acute Hospitals - Acute Care Bed Count",
          legend = list(x = 0.05, y = 0.95),
         yaxis = list(title = "Number of Beds", tickformat = ".0f"),
         xaxis = list(title = "Source: San Francisco Department of Public Health"),
         hovermode = "compare")
```

```{r}
covid19sf_hospital %>%
  filter(bed_type == "Intensive Care") %>%
  mutate(status = factor(status, levels = c("Other Patients", "COVID-19 (Confirmed & Suspected)", "Available"))) %>%
  plot_ly(x = ~ date,
          y = ~ count,
          type = 'scatter', 
          mode = 'none', 
          color = ~status,
          stackgroup = 'one') %>%
    layout(title = "SF Acute Hospitals - Intensive Care Bed Count",
          legend = list(x = 0.05, y = 1),
         yaxis = list(title = "Number of Beds", tickformat = ".0f"),
         xaxis = list(title = "Source: San Francisco Department of Public Health"),
         hovermode = "compare")
```


### Covid19 Hospitalizations 

The `covid19sf_hospitalizations` provides daily snapshot of the number of patients in San Francisco hospitals by hospitalization unit and Covid19 status using the following variables:

* `dphcategory` - the hospital unit type, either `ICU` (intensive care unit) or `Med/Surg` (acute care).
* `covidstatus` - the Covid19 status, either `COVID+` (confirmed cases) or `PUI` (patient under investigation)

The below example summarised the hospitalizations status as of the most recent date in the data:


```{r}
data(covid19sf_hospitalizations)

covid19sf_hospitalizations %>%
  filter(reportdate == max(reportdate)) %>%
  group_by(dphcategory, covidstatus) %>%
  summarise(total = sum(patientcount))
```



