---
title: "Introduction to the covid19sf Datasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(covid19sf)
```

The covid19sf package provides a daily summary of the covid19 cases in San Francisco. The package includes the following datasets:


* `covid19sf_geo` -  Confirmed cases and deaths summarized by geography
* `covid19sf_hospital` -  Hospital capacity data
* `covid19sf_hospitalizations` -  Hospitalizations data
* `covid19sf_housing` -  Alternative housing sites
* `covid19sf_test_loc` -  Testing locations
* `covid19sf_tests` -  Daily number of tests
* `covid19sf_vaccine_demo` -  Summary of vaccine doses given to San Franciscans by demographics groups (age and race)
* `covid19sf_vaccine_demo_ts` -  Time series view of vaccine doses given to San Franciscans by demographics groups (age and race)
* `covid19sf_vaccine_geo` - COVID-19 vaccines given to San Franciscans by geography
* `covid19sf_population` - COVID-19 cases by population characteristics over time

The following dataset were deprecated and replaced by the `covid19sf_population` dataset:

* `covid19sf_demo`- Cases summarized by date, transmission and case disposition
* `covid19sf_homeless` -  Confirmed cases by homelessness
* `covid19sf_age` -  Cases summarized by age group
* `covid19sf_gender` -  Confirmed cases summarized by gender
* `covid19sf_summary` -  Cases summarized by date, transmission and case disposition

**Data soucre:** San Francisco, Department of Public Health - Population Health Division through the San Francisco [Opne Data protal website](https://datasf.org/opendata/)

