---
title: "San Francisco Vaccination Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vaccine_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message=FALSE, 
  warning=FALSE,
  collapse = TRUE,
  fig.height=5, 
  fig.width=8,
  comment = "#>"
)
```

The covid19sf_vaccine_demo and covid19sf_vaccine_demo_ts datasets provide demographics information on the San Francisco vaccination effort:

-   `covid19sf_vaccine_demo` - Summary of vaccine doses given to San Franciscans by demographics groups (age and race)
-   `covid19sf_vaccine_demo_ts` - Time series view of vaccine doses given to San Franciscans by demographics groups (age and race)

### San Francisco vaccination summary

The `covid19sf_vaccine_demo` summarizes the COVID-19 vaccine doses given to San Franciscans by demographics (age and race groups) and includes the following fields:

-   `overall_segment` - Segment (universe) of analysis. Unique combination of administering_provider_type, age_group, and demographic_group. Filter to a single option to derive meaningful totals.
-   `administering_provider_type` - Providers included in a given overall_segment. Two possible values: 'All' (including SF DPH) or 'DPH Only'
-   `age_group` - Age range included in a given overall_segment
-   `demographic_group` - Type of demographic group included in a given overall_segment (e.g. Age, Race/Ethnicity)
-   `demographic_subgroup` - Specific demographic group counted in a given record (e.g. 16-24, Asian)
-   `demographic_subgroup_sort_order` - Numeric sort order for all demographic_subgroups. Convenient for maintaining consistent ordering across multiple data visualizations.
-   `total_1st_doses` - Total number of first doses administered
-   `total_2nd_doses` - Total number of second doses administered
-   `total_single_doses` - Total number of single dose vaccines administered
-   `total_recipients` - Total number of unique vaccine recipients
-   `total_series_completed` - Total number of individuals fully vaccinated (those having received the second dose of a two-dose vaccine or one dose of a single-dose vaccine)
-   `subgroup_population` - 2018 5-year American Community Survey population estimates for given demographic_subgroup
-   `age_group_population` - 2018 5-year American Community Survey population estimates for overall age_group\
-   `data_as_of` - Timestamp for last update date in source system
-   `data_loaded_at` - Timestamp when the record (row) was most recently updated in Socrata

```{r setup}
library(covid19sf)

data("covid19sf_vaccine_demo")

head(covid19sf_vaccine_demo)

str(covid19sf_vaccine_demo)
```

Using this dataset, we can answer the following questions:

-   What is the proportion of San Francisco population that got vaccinated?
-   How many people got two doses of the vaccine (e.g., Pfizer or Moderna) vs single dose (J&J)?
-   What is the distribution of the vaccine doses given to San Franciscans by demographics groups (age and race)?

#### Doses given to San Franciscans by age group:

In the following example, we will explore the distribution of San Francisco population by age group. Let's start by filtering the data by all providers and age group demographics:

```{r }
library(dplyr)
library(plotly)

df_age <- covid19sf_vaccine_demo %>%
  filter(administering_provider_type == "All Providers", 
         demographic_group == "Age Bracket",
         age_group == "All")

df_age
```

Next, we will sort the data by age group and convert the `demographic_subgroup` into an ordered factor:

```{r}
df_age <- df_age %>%
  arrange(demographic_subgroup_sort_order) %>%
  mutate(age_ordered = factor(demographic_subgroup, levels = demographic_subgroup))
```

Let's start by plotting the distribution of San Francisco vaccinated population using the following groups:

-   **Two Doses** - San Franciscans who received the two doses of either Pfizer or Moderna vaccine
-   **Single Dose** - San Franciscans who received the J&J single-dose vaccine
-   **First Dose** - San Franciscans who received the first dose out of two of either Pfizer or Moderna vaccine
-   **Not Vaccinated** - San Franciscans who did not receive any COVID-19 vaccine

Since the `total_2nd_doses` variable (received the second dose) is a subset of `total_1st_doses` (received only the first out of two doses), we will have to exclude it:

```{r}
df_age <- df_age %>%
  mutate(only_1st_dose = total_1st_doses - total_2nd_doses)
```

Next, we will collapse and aggregate the data by vaccination status:

```{r}
df_summary <- df_age %>%
  summarise(`First Dose` = sum(only_1st_dose),
            `Two Doses` = sum(total_2nd_doses),
            `Single Dose` = sum(total_single_doses),
            `Total Pop` = sum(subgroup_population)) %>%
  mutate(`Not Received Vaccine` = `Total Pop` - `First Dose` - `Two Doses` - `Single Dose`) %>%
  select(- `Total Pop`) %>%
  t() %>%
  as.data.frame() %>%
  select(count = V1)

df_summary$type <- rownames(df_summary)

head(df_summary)
```

Using the `df_summary`, let's plot the distribution of San Francisco vaccination status:

```{r}
plot_ly(data = df_summary,
        labels = ~ type,
        values = ~ count,
        textposition = 'inside',
        textinfo = 'label+percent',
        insidetextfont = list(color = '#FFFFFF',
                              size = 16),
        marker = list(colors = c("rgb(35, 90, 122)", "rgb(128, 200, 215)",
                                 "rgb(74, 168, 192)", "rgb(225, 100, 74)"),
                      line = list(color = '#FFFFFF', width = 1)),
        type = "pie",
        showlegend = FALSE) %>%
  layout(title = "Distribution of COVID-19 Vaccine Doses Given to San Franciscans",
         font = list(family = "Arieal",
                     size = 18),
         margin = list(t = 80, b = 60)) %>%
  add_annotations(text = paste("- Including children under the age of 12",
                               "- The First and Second Doses refer to either Pfizer or Moderna vaccine",
                               "- Single Dose refers to J&J vaccine", sep = "<br>"),
                  font = list(family = "Arieal",
                     size = 14),
                  
                  x = -0.02,
                  y = -0.15,
                  align = "left",
                  xref = "paper",
                  yref = "paper",
                  showarrow = FALSE)
```

As can see in the plot above, about `r paste(round(100 *sum(df_summary[which(df_summary$type%in% c("Single Dose", "Two Doses")),]$count) / sum(df_summary$count)),"%", sep = "")` of the SF population completed the vaccination process, and `r paste(round(100 *sum(df_summary[which(df_summary$type%in% c("First Dose")),]$count) / sum(df_summary$count)),"%", sep = "")` received the first dose out of two. About `r paste(round(100 *sum(df_summary[which(df_summary$type%in% c("Not Received Vaccine")),]$count) / sum(df_summary$count)),"%", sep = "")` did not vaccinate. Note that the distribution includes children under the age of 12, which at this point are not vaccinated. If we want to get a better distribution of the vaccination effort of the SF population, let's exclude children under the age of 12:

```{r}
df_summary2 <- df_age %>%
  filter(demographic_subgroup != "Under 12") %>%
  summarise(`First Dose` = sum(only_1st_dose),
            `Two Doses` = sum(total_2nd_doses),
            `Single Dose` = sum(total_single_doses),
            `Total Pop` = sum(subgroup_population)) %>%
  mutate(`Not Received Vaccine` = `Total Pop` - `First Dose` - `Two Doses` - `Single Dose`) %>%
  select(- `Total Pop`) %>%
  t() %>%
  as.data.frame() %>%
  select(count = V1)

df_summary2$type <- rownames(df_summary2)

plot_ly(data = df_summary2,
        labels = ~ type,
        values = ~ count,
        textposition = 'inside',
        textinfo = 'label+percent',
        insidetextfont = list(color = '#FFFFFF',
                              size = 16),
        marker = list(colors = c("rgb(35, 90, 122)", "rgb(128, 200, 215)",
                                 "rgb(74, 168, 192)", "rgb(225, 100, 74)"),
                      line = list(color = '#FFFFFF', width = 1)),
        type = "pie",
        showlegend = FALSE) %>%
  layout(title = "Distribution of COVID-19 Vaccine Doses Given to San Franciscans",
         font = list(family = "Arieal",
                     size = 18),
         margin = list(t = 80, b = 60)) %>%
  add_annotations(text = paste("- Excluding children under the age of 12",
                               "- The First and Second Doses refer to either Pfizer or Moderna vaccine",
                               "- Single Dose refers to J&J vaccine", sep = "<br>"),
                  font = list(family = "Arieal",
                     size = 14),
                  
                  x = -0.02,
                  y = -0.15,
                  align = "left",
                  xref = "paper",
                  yref = "paper",
                  showarrow = FALSE)
```

After excluding children under the age of 12 group, we can see from the plot above that `r paste(round(100 *sum(df_summary2[which(df_summary2$type%in% c("Single Dose", "Two Doses")),]$count) / sum(df_summary2$count)),"%", sep = "")` of the SF population (age 12 and above) completed the vaccination process, and `r paste(round(100 *sum(df_summary2[which(df_summary2$type%in% c("First Dose")),]$count) / sum(df_summary2$count)),"%", sep = "")` received the first dose.

Let's now plot the total population that received the full doses of the vaccine (single for J&J and two for Pfizer and Moderna):

```{r }
plot_ly(data = df_age) %>%
  add_trace(x = ~ age_ordered,
            y = ~ total_2nd_doses,
            type = "bar",
            name = "Pfizer or Moderna") %>%
  add_trace(x = ~ age_ordered,
            type = "bar",
            y = ~ total_single_doses,
            name = "J&J") %>%
  layout(title = "San Francisco Fully Vaccinated Population by Age and Vaccine Type",
         barmode = "stack",
         yaxis = list(title = "Population"),
         xaxis = list(title = "Age Group"))
```

Last but not least, we will use a bar plot to describe the distribution, in percentage, between the city population that fully, partial, and not vaccinated. We will pre-calculate the percentage break down for each age group:

```{r}

d1a <- covid19sf_vaccine_demo %>%
  filter(administering_provider_type == "All Providers",
                demographic_group == "Age Bracket",
                age_group == "All") %>%
  mutate(only_1st = total_1st_doses - total_2nd_doses,
                not_vaccinated = subgroup_population - total_series_completed - only_1st,
                first_per = 100 * only_1st / subgroup_population,
                second_per = 100 * total_2nd_doses / subgroup_population,
                single_per = 100 * total_single_doses / subgroup_population,
                not_vaccinated_per = 100 * not_vaccinated / subgroup_population,
                complete_per = 100 - not_vaccinated_per - first_per) %>%
  arrange(demographic_subgroup_sort_order) %>%
  mutate(age = factor(demographic_subgroup, levels = demographic_subgroup)) %>%
  select(age, first_per, second_per, single_per, not_vaccinated_per, complete_per)


d1a

```

Inspired from The Economist data visualization, for the following plot, we will use The Economist default color palette for bar plot:

```{r}
# plot setting
font_size <- 22
opacity <- 0.95
background <- "rgb(225, 236, 242)"
complete_color <- paste("rgba(47, 106, 158,", opacity ,")", sep = "")
first_color <-  paste("rgba(100, 184, 206,", opacity ,")", sep = "")
not_vaccinated_color <- paste("rgba(162, 32, 61,", opacity ,")", sep = "")

# Plot
plot_ly(data = d1a,
        x = ~ complete_per,
        y = ~ age,
        type = 'bar', orientation = 'h',
        name = "Completed Two/Single Doses",
        marker = list(color = complete_color,
                      line = list(color = toRGB("gray50"), width = 1))) %>%
  add_trace(x = ~ first_per ,
            marker = list(color =first_color),
            name = "First Dose") %>%
  add_trace(x = ~ not_vaccinated_per ,
            marker = list(color = not_vaccinated_color),
            name = "Not Vaccinated") %>%
  layout(xaxis = list(title = "",
                      showgrid = TRUE,
                      showline = FALSE,
                      showticklabels = TRUE,
                      zeroline = TRUE,
                      tickwidth = 2,
                      tickcolor = toRGB("gray50"),
                      gridwidth = 2,
                      gridcolor = toRGB("gray50"),
                      side ="top",
                      # domain = c(0.1, 1),
                      ticksuffix = "%"),
         yaxis = list(title = "Age Group",
                      # domain = c(0, 0.8),
                      showgrid = FALSE,
                      showline = FALSE,
                      showticklabels = TRUE,
                      zeroline = FALSE),
         barmode = "stack",
         plot_bgcolor = background,
         paper_bgcolor = background,
         margin = list(l = 70, t = 100, b = 60),
         legend = list(orientation = 'h')) %>%
  add_annotations(text = "Source: San Francisco, Department of Public Health - Population Health Division",
                  font = list(size = 16,
                              family = "Ariel"),
                  align = "left",
                  x = 0,
                  y = - 0.08,
                  xref = "paper",
                  yref = "paper",
                  showarrow = FALSE) %>%
  add_annotations(text = "San Francisco COVID-19 Vaccination Distribution by Age Group",
                  y = 1.19,
                  x = -0.03,
                  font = list(family = "Ariel",
                              size =  24,
                              color = "black"),
                  yref = "paper",
                  xref = "paper",
                  align = "left",
                  valign = "middle",
                  showarrow = FALSE) 
```



### San Francisco vaccination time series

The `covid19sf_vaccine_demo_ts` dataset is a time series format of the `covid19sf_vaccine_demo` dataset, and includes the following fields:

* `date_administered` - Date vaccination administered
* `overall_segment` -Segment (universe) of analysis.
   Unique combination of administering_provider_type, age_group, and demographic_group.
   Filter to a single option to derive meaningful totals.
* `administering_provider_type` - Providers included in a given overall_segment. Two possible values: 'All' (including SF DPH) or 'DPH Only'
* `age_group` - Age range included in a given overall_segment
* `demographic_group` - Type of demographic group included in a given overall_segment (e.g. Age, Race/Ethnicity)
* `demographic_subgroup` - Specific demographic group counted in a given record (e.g. 16-24, Asian)
* `demographic_subgroup_sort_order` - Numeric sort order for all demographic_subgroup. Convenient for maintaining consistent ordering across multiple data visualizations.
* `new_1st_doses` - Count of 1st doses administered for vaccines that take two doses to complete
* `new_2nd_doses` - Count of 2nd doses administered for vaccines that take two doses to complete
* `new_single_doses` - Count of doses administered for vaccines that take one dose to complete
* `new_series_completed` - Count of individuals newly fully vaccinated on a given day (given the 2nd dose of a two-dose vaccine or one dose of a single dose vaccine)
* `new_recipients` - Count of individuals vaccinated (with any dose) for the first time according to CA's records
* `cumulative_1st_doses` - Cumulative total of 1st doses administered for vaccines that take two doses to complete
* `cumulative_2nd_doses` - Cumulative total of 2nd doses administered for vaccines that take two doses to complete
* `cumulative_single_doses` - Cumulative total of doses administered for vaccines that take one dose to complete
* `cumulative_series_completed` - Cumulative total individuals fully vaccinated (given the 2nd dose of a two-dose vaccine or one dose of a single dose vaccine)
* `cumulative_recipients` - Cumulative total individuals vaccinated (with any dose) according to CA's records
* `subgroup_population` - American Community Survey population estimates for given demographic_subgroup
* `age_group_population` - American Community Survey population estimates for overall age_group
* `data_as_of` - Timestamp for last update date in source system
* `data_loaded_at` - Timestamp when the record (row) was most recently updated here in the Open Data Portal

Let's load and review the dataset:

```{r}
data("covid19sf_vaccine_demo_ts")

str(covid19sf_vaccine_demo_ts)

unique(covid19sf_vaccine_demo_ts$demographic_group)
```

This dataset enables us to review the vaccination progress of SF population by age and race/ethnicity group over time. Let's review the vaccination of the SF population by race/ethnicity group:

```{r}
df <- covid19sf_vaccine_demo_ts %>%
  dplyr::filter(overall_segment == "All Ages by Race/Ethnicity, Administered by All Providers") 

table(df$demographic_subgroup)

df %>%
  plot_ly(x = ~ date_administered,
          y = ~ new_series_completed,
          color = ~ demographic_subgroup,
          type = "scatter",
          mode = "line") %>%
  layout(title = "San Francisco COVID-19 Daily Vaccination by Race/Ethnicity",
         yaxis = list(title = "Population"),
         xaxis = list(title = "Source: San Francisco, Department of Public Health - Population Health Division"),
         legend = list(x = 0, y = 1.1),
         margin = list(t = 90))
```

Similarly, we can plot the vaccination progress over time by race/ethnicity group. Note that there is some calculation error for `Other Race` demographic sub-group as the cumulative values of the group surpass the sub-group population. Therefore, we will exclude it from the calculation: 

```{r}
df %>%
  filter(demographic_subgroup != "Other Race") %>%
  plot_ly(x = ~ date_administered,
                  y = ~ 100 *cumulative_series_completed /subgroup_population,
                  color = ~ demographic_subgroup,
                  hoverinfo = "text",
                  text = ~ paste(demographic_subgroup, "<br>",
                                 "Date:", date_administered, "<br>",
                                 "Sub-Group Population:", subgroup_population, "<br>",
                                 "Completed Series:", cumulative_series_completed,
                                 paste("(", round(100 *cumulative_series_completed /subgroup_population) ,"%)", sep = "")),
                  type = "scatter",
                  mode = "line") %>%
  layout(title = "Proportion of San Francisco COVID-19 Vaccinated Population by Race/Ethnicity",
         yaxis = list(title = "Percentage", 
                      ticksuffix = "%"),
         xaxis = list(title = "Source: San Francisco, Department of Public Health - Population Health Division"),
         legend = list(x = 0.05, y = 0.95),
         margin = list(t = 90))
```

